# Course Project 1 (Reproducible Research)
## Submitted by Kritika Gulati

### 1. Loading and Preprocessing the data

* Loading the data
* Processing/transforming the data into a format suitable for analysis

```{r, echo=TRUE, cache=TRUE}
x<-read.csv("activity.csv", header=TRUE, sep = ",")
x$date<-as.Date(x$date)
```

### 2. What is the mean total number of steps taken per day?

* Calculating the total number of steps per day

```{r, echo=TRUE, cache=TRUE}
y<-NULL
for (i in 1:length(unique(x$date))) {
  d1<-subset(x, x$date==unique(x$date)[i])
  y[i]<-sum(d1$steps, na.rm = TRUE)
}
d2<-data.frame(date=unique(x$date), Steps=y)
d2
```

* Histogram of the total number of steps taken each day

```{r, echo=TRUE, cache=TRUE}
hist(d2$Steps, xlab = "Total number of steps taken per day", 
     main = "Histogram of total number of steps taken each day", 
     col = "light pink")
```


* The mean and median of the total number of steps taken per day

```{r, echo=TRUE, cache=TRUE}
print(paste("Mean of the total number of steps taken per day:", mean(y)))
print(paste("Median of the total number of steps taken per day:", median(y)))
```

### 3. Average daily activity pattern

* Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE, cache=TRUE}
y<-0
a<-NULL
a1<-nrow(subset(x,x$date==unique(x$date)[1]))
for (i in 1:a1) {
  j<-i
  while (j<=nrow(x)){
    y[i]<-sum(y[i], x$steps[j], na.rm = TRUE)
  j<-j+a1 }
  a[i]<-y[i]/a1
}

plot(unique(x$interval), a, type = "l", xlab = "5 minute Interval", 
     ylab = "Average number of steps taken across all days", 
     main = "time series plot")
```

* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE, cache=FALSE}
res1<-which(a==max(a))
res3<-unique(x$interval)[res1]
res2<-max(a)
```

Hence, `r res3` is the 5-minute interval that contains the maximum number 
of steps, `r res2` on average.

### 4. Imputing missing values

* Calculating the total number of missing values in the dataset (i.e total
number of rows with NAs)

```{r, echo=TRUE, cache=TRUE}
k<-0
for(i in 1:nrow(x)){
  if(is.na(x$steps[i])==TRUE)
    k<-k+1
}
```

There are `r k` missing values in the dataset. 

* A strategy for filling in the missing values in the dataset.
We will be filling in the missing values for a particular day in the dataset with the mean of the number of steps taken in that day.

*  A new dataset that is equal to the original dataset but with the missing data filled in

```{r, echo=TRUE, cache=TRUE}
y<-NULL
x1<-x
for (i in 1:length(unique(x1$date))) {
  d1<-subset(x1, x1$date==unique(x1$date)[i])
  y[i]<-mean(d1$steps, na.rm = TRUE)
}
m<-which(is.nan(y)==TRUE)
for (i in m) {
  y[i]<-0
}
i<-1
j<-1
while (i<=nrow(x1)) { 
    for (k in 1:a1) {
      if(is.na(x1$steps[k+i-1])==TRUE)
        x1$steps[k+i-1]<-y[j]
    }
  j<-j+1
   i<-i+a1 
}

```

*  Histogram of the total number of steps taken each day with mean and median total number of steps taken per day

```{r, echo=TRUE, cache=TRUE}
y<-NULL
for (i in 1:length(unique(x1$date))) {
  d1<-subset(x1, x1$date==unique(x1$date)[i])
  y[i]<-sum(d1$steps)
}
d2<-data.frame(date=unique(x1$date), Steps=y)

hist(d2$Steps, xlab = "Total number of steps taken per day", 
     main = "Histogram of total number of steps taken each day", 
     col = "pink")
print(paste("Mean of the total number of steps taken per day:", mean(y)))
print(paste("Median of the total number of steps taken per day:", median(y)))

```

Hence, we see that the mean and median values stay the same as the first part of the assignment. Hence, there is no impact of imputing missing values in the dataset. 

### 5. Are there differences in activity patterns between weekdays and weekends?

* Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE, cache=TRUE}
x1<-cbind(x1, day=weekdays(x1$date))
for(i in 1:nrow(x)){
  if(x1$day[i]=="Saturday" | x1$day[i]=="Sunday")
    x1$day[i]<-"weekend"
  else
    x1$day[i]<-"weekday"
}
x1$day<-as.factor(x1$day)

```

* A panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo=TRUE, cache=TRUE}
x2<-subset(x1, x1$day=="weekday")
y<-0
a2<-NULL
a1<-nrow(subset(x2,x2$date==unique(x2$date)[1]))
for (i in 1:a1) {
  j<-i
  while (j<=nrow(x2)){
    y[i]<-sum(y[i], x2$steps[j], na.rm = TRUE)
    j<-j+a1 }
  a2[i]<-y[i]/a1
}
x3<-subset(x1, x1$day=="weekend")
y<-0
a3<-NULL
a1<-nrow(subset(x3,x3$date==unique(x3$date)[1]))
for (i in 1:a1) {
  j<-i
  while (j<=nrow(x3)){
    y[i]<-sum(y[i], x3$steps[j], na.rm = TRUE)
    j<-j+a1 }
  a3[i]<-y[i]/a1
}
par(mfcol=c(1,2), mar=c(4,4,1,1))
plot(unique(x1$interval), a2, type="l", xlab ="Interval", 
     ylab="Average steps taken")
legend("topright", legend = "Weekday")
plot(unique(x1$interval), a3, type="l", ylim = c(0,30), xlab = "Interval",
     ylab = "Average Steps taken")
legend("topright", legend="Weekend")
```

